A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-06-12 manuscript.qmd
various formatting improvements to manuscript.qmd

---
title: "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle: "Simulation-Based Sample Size Planning for GLMMs"
author:
  - name: Florian Pargent
    equal-contributor: true
    orcid: 0000-0002-2388-553X
    roles:
      - formal analysis
      - methodology
      - visualization
      - writing
      - editing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
  - name: Timo K. Koch
    corresponding: true
    equal-contributor: true
    orcid: 0000-0001-6728-2063
    email: timo.koch@unisg.ch
    roles:
      - conceptualization
      - formal analysis
      - writing
      - editing
    affiliations:
      - name: University of St. Gallen 
        department: Institute of Behavioral Science & Technology
        address: Torstrasse 25
        city: St. Gallen
        country: Switzerland
        postal-code: 9000
      - ref: lmu
  - name: Anne-Kathrin Kleine
    roles:
      - formal analysis
      - editing
    affiliations:
      - ref: lmu
  - name: Eva Lermer
    roles:
      - conceptualization
      - funding aquisition
      - editing
    affiliations:
      - ref: lmu
      - name: Technical University of Applied Sciences Augsburg
        department: Department of Business Psychology
  - name: Susanne Gaube
    roles:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: lmu
      - name: University College London
        department: Global Business School for Health
author-note:
  disclosures:
    data-sharing: "All materials (reproducible manuscript incl. R code, simulation results) are available in the project's repository on the Open Science Framework (OSF) at https://osf.io/dhwf4/"
    conflict-of-interest: "The authors declare that there were no conflicts of interest with respect to the authorship or the publication of this article."
    financial-support: "This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525)."
    gratitude: ~
    authorship-agreements: "Florian Pargent and Timo K. Koch contributed equally to this work."
abstract: "When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. The recent replication crisis underlines the need for more rigorous statistical methodology and adequately powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. However, available methods and software for a priori sample size planning for GLMMs are often limited to specific designs. Tailored data simulation approaches are a more flexible alternative. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori sample size planning with GLMMs. We not only focus on power analysis but also explain how to use the precision of parameter estimates to determine appropriate sample sizes. We conclude with an outlook on the increasing importance of simulation-based sample size planning."
keywords: [power analysis, precision, data simulation, sample size planning, generalized linear mixed model]
bibliography: [Simulation Tutorial Paper.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

```{r}
#| label: setup
#| include: false
library("papaja")
library("tidyverse")
library("faux")
library("lme4")
library("multcomp")
library("marginaleffects")
library("future")
library("furrr")
library("binom")
r_refs("r-references.bib")
```

@@

Performing tailored simulation-based sample size planning is more complicated and time-consuming than using the existing software tools outlined in TABLE XXX. The most important circumstances under which tailored simulation-based sample size planning is necessary are 1) complex study designs, 2) complex statistical hypotheses, 3) planning for precision, 4) no available prior studies or pilot data.

@@

Although planning for precision [@lakensSampleSizeJustification2022; @maxwellSampleSizePlanning2008; @cummingNewStatisticsWhy2014] has been increasingly discussed as a useful strategy for empirical research, all available software packages for sample size planning with GLMMs are based on power analysis.
Therefore, researchers that want to want to apply an estimation strategy in their studies, without testing any statistical hypotheses, currently cannot use the software packages outlined in TABLE XXX.
<!-- As long as one is interested only in confidence intervals of a single model parameter, it might be possible to adapt existing software packages to plan for precision. However, we demonstrate in our case study that in realistic research projects with GLMMs, the contrasts of interests and their respective confidence intervals are usually situated on the scale of the (expected) outcome in contrast to the scale of the model parameters.  -->
However, tailored simulation-based sample size planning can easily handle the planning for precision approach [@maxwellSampleSizePlanning2008]. The only change in procedure is that instead of computing HTs for each simulated dataset and estimating statistical power across repetitions, CIs are computed for each simulated dataset, and the expected width is estimated.

@@

Although the details differ depending on the specific study characteristics, each tailored simulation-based sample size planning requires a series of steps and decisions. We will introduce each step in a theoretical section, followed by the practical application based on a hypothetical case study. All code in this manuscript and simulation results are available in the project’s repository on the Open Science Framework (https://osf.io/dhwf4/).

@@

<!-- When the empirical estimand has been specified, the last step is to define how a concrete estimate of the empirical estimand shall be computed based on the observed study data. When a regression model is used for estimation, this includes decisions like which statistical framework estimation technique and software package are used to fit the model and how to compute HTs or CIs. -->

@@

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, average student, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{wrong advice, average student, average scan})
\end{aligned}
$$
For example, the first expression is concerned with the difference between the probability that a correct diagnosis is made if correct AI-advice is presented and the probability that a correct diagnosis is made if no AI-advice is presented. This contrast is quantified for a hypothetical *typical* expert and a *typical* scan, where *typical* is usually defined as an average score on all attributes of the expert or scan.^[Note that a different estimand would be the so-called average treatment effect (ATE). For the ATE, the probability contrast is defined for each combination of expert and scan, and then these contrasts are averaged across all experts and scans from the target population [@lundbergWhatYourEstimand2021]].
To complete our definition of the estimand, we have to define our target population that consists of persons, stimuli, and AI-advice: With respect to persons, we are only interested in expert radiologists and medical students at German universities. With respect to stimuli, we are only interested in the head CT scans made from subjects that do or do not suffer from intracerebral hemorrhage. Lastly, we are only interested in AI-advice given by a specific AI-enabled diagnostic decision support system.

@@

The discussed inequality of conditional and marginal effects in GLMMs [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult.
One must be careful when specifying parameter values based on previous studies or pilot data that use the marginal interpretation (e.g., a pilot study providing an estimate of how often radiologists make an accurate diagnosis based on brain scans). 
However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge.
Then, we can iteratively adjust the parameter values until we are satisfied with the implied distributions.

In the last section, we simulated a large dataset and computed descriptive statistics, the relative frequencies of correct diagnoses, for each experimental condition.
We will now use the model implied probability of each simulated data point (stored in the variable `y_prob`) to visualize the whole model implied marginal distribution of correct diagnoses for each experimental condition.

```{r}
#| label: fig-margdist1
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions."
#| echo: true
library(ggdist)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@fig-margdist1 shows the model implied marginal distributions, including the mean, 66% and 95% intervals. We can see that, indeed, the average probabilities (black dots) slightly differ from the probabilities of average subjects and items considered in the previous section. This difference increases with the variability of the random effects.

We can use plots like the one above as a useful tool to decide whether the specified standard deviations of the subject and item random intercepts ($\sigma_S$ and $\sigma_I$) are reasonable by comparing the ranges and overlap between conditions to domain knowledge.

In the next plot, we have set the item standard deviation to almost zero ($\sigma_I = 0.01$). This gives us a better way to see the variability between persons.

```{r}
#| label: fig-margdist2
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of item random intercepts to 0.01."
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0i = 0.01)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

As an example, @fig-margdist2 reveals a number of implicit assumptions about the comparison between experts and non-experts: With wrong advice, virtually all experts have a higher probability of making a correct diagnosis compared to non-experts when considering only items with average difficulty. In contrast, there is considerable overlap in probability between experts and non-experts with no advice and even higher overlap with correct advice. Patterns like these should be considered carefully and discussed with the domain experts. Parameter values ($\beta$ parameters, and $\sigma_S$) should be adjusted if the implications do not seem reasonable.
We could also have a closer look at variability between items by setting the subject standard deviation to almost zero ($\sigma_S = 0.01$).

```{r}
#| include: false
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 0.01)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

The final plot demonstrates that these plots are also useful for spotting standard deviations that are specified too high. For @fig-margdist3 we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are usually close to either 0 or 1, which is not a plausible assumption. These high standard deviations do not account for the inherent variability and complexity of human performance. For example, we would expect that a participant with low ability compared to other task experts to solve a difficult item with a probability substantially larger than zero even when presented with wrong advice.

```{r}
#| label: fig-margdist3
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of subject and item random intercepts to 3."
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 3, sd_u0i = 3)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@@

<!-- The lme4 package includes a function called `simulate` that allows researchers to simulate the dependent variable based on the same model formula used for model fitting, enabling simulation-based power analyses and other related analyses.  -->

<!-- [^1]: For Bayesian GLMMs, the brms R package is currently the most prominent option [@burknerBrmsPackageBayesian2017]. -->

<!-- However, the model parameterization used by the lme4 package is quite technical, making it difficult for applied researchers to determine whether their specified population model (i.e., the theoretical model that describes the underlying data generation process for a specific population of interest) implies plausible associations in their simulated data. Therefore, for our case study, we have written our own function to simulate data for GLMMs from first principles (i.e., creating synthetic data step by step instead of using black box functions). -->
<!-- We think this choice assists applied researchers in better understanding all model assumptions before using lme4 to analyze the simulated datasets.[^2] -->

<!-- [^2]: A less flexible alternative would be to use the simr package [@greenSIMRPackagePower2016], which can be used to both simulate data and perform power analysis for models supported by the lme4 package. -->

In our case study, we use the lme4 R package [@batesFittingLinearMixedEffects2015], which is a state-of-the-art tool for fitting frequentist GLMMs.[^1]
For the current example, we simulate data according to our model, in which 100 subjects respond to 50 items (we use `set.seed` to make the simulation reproducible). However, for the sake of the exercise, we can imagine that this would be real data resulting from our future experiment and think about how we would analyze this data.

@@

<!-- $$ -->
<!-- \begin{aligned} -->
<!-- & \text{omv}[\beta_0 + \beta_a \cdot 1 + \beta_c \cdot 1 + \beta_e \cdot 1 + \beta_{ea} \cdot 1 \cdot 1 + \beta_{ec} \cdot 1 \cdot 1]\\ -->
<!-- & \quad - \text{inverse_logit}[\beta_0 + \beta_a \cdot 0 + \beta_c \cdot 0 + \beta_e \cdot 1 + \beta_{ea} \cdot 1 \cdot 0 + \beta_{ec} \cdot 1 \cdot 0] = \\ -->
<!-- & = \text{inverse_logit}[\beta_0 + \beta_a  + \beta_c + \beta_e + \beta_{ea} + \beta_{ec} ]\\ -->
<!-- & \quad - \text{inverse_logit}[\beta_0 + \beta_e ] \\ -->
<!-- \end{aligned} -->
<!-- $$ -->

We could use our knowledge of the structure of the GLMM to determine the exact formula needed to compute the contrasts of interest and then plug in the parameter estimates manually from the `summary(fit)` output.
However, this would be tedious and we can use R to compute this contrast without doing the math.

Using the first contrast as our example, we could apply the `predict` function of the lme4 package to compute the predicted probability for a correct diagnosis based on our fitted model, plug in the two sets of predictor values, and compute the difference between the two estimated probabilities.

```{r}
#| echo: true
pred_df <- data.frame(advice_present = c(1, 0), advice_correct = c(1, 0), 
  expert = c(1, 1))
pred_df
preds <- predict(fit, newdata = pred_df, type = "response", re.form = NA)
preds
preds[1] - preds[2]
```

The argument `type = "response"` specifies that predictions are made on the probability scale (instead of the log-odds scale of the $\beta$ parameters), while `re.form = NA` sets all random effects to 0.
We could use this method to compute point estimates for all four contrasts that are part of our estimand.
However, depending on whether we are interested in hypothesis testing or parameter estimation, we also need a method to compute HTs or CI.

The marginaleffects package [@R-marginaleffects] is a very flexible, increasingly popular package to compute HTs and CIs for contrasts with a variety of statistical models, including GLMMs estimated with lme4.
First, we specify a grid of all combinations of predictor variable and then compute estimated probabilities for all experimental conditions in our experiment with the `predictions` function:

```{r}
#| echo: true
library(tidyverse)
library(marginaleffects)
pred_df <- expand_grid(advice_present = 0:1, advice_correct = 0:1, expert = 0:1)
pred_df
probs <- predictions(fit, newdata = pred_df, type = "response", re.form = NA)
probs
```

@@

```{r}
#| echo: true
probs |> 
  hypotheses(hypothesis = c(
    "b8 = b2",
    "b2 = b6",
    "b7 = b1",
    "b1 = b5"),
    equivalence = c(0, 0))
```

@@

If we chose hypothesis testing for our case study, we would test a combined null hypothesis $H_0$ that consists of four separate null hypotheses:

$$
\begin{aligned}
H_{01}:\ & P(Y=1|advice\_present = 1, advice\_correct = 1, expert = 1, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \\
H_{02}:\ &P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 1, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \\
H_{03}:\ &P(Y=1|advice\_present = 1, advice\_correct = 1, expert = 0, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) \\
H_{04}:\ & P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 1, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0)
\end{aligned}
$$
The combined null hypothesis $H_0$ should only be rejected if **all** individual null hypotheses are rejected [i.e., intersection-union setting\; @dmitrienkoTraditionalMultiplicityAdjustment2013].
In such cases, the error probabilities do not accumulate, and we would waste power when correcting for multiple tests.

@@

<!-- Due to this complicated relationship, we argue not to focus too much on interpreting single model parameters when working with GLMMs. Instead, it can be more intuitive to consider model predictions and the model-implied distribution of the dependent variable for each experimental condition across all subjects and items. -->

<!-- With the marginaleffects package [@R-marginaleffects], we can easily compute predictions for all observations in the dataset based on the fitted GLMM (including all fixed **and** random effects), and plot the average probability with confidence intervals for each experimental condition in Figure \@ref(fig:marginaleffects): -->

<!-- (ref:marginaleffects) Marginal distributions including means and 95% confidence intervals for all experimental conditions computed with the marginaleffects package. -->

<!-- ```{r marginaleffects, message=FALSE, fig.cap="(ref:marginaleffects)"} -->
<!-- library(marginaleffects) -->
<!-- plot_predictions(fit, by = c("advice_present", "advice_correct", "expert"), -->
<!--   type = "response") + ylim(c(0.3, 1)) -->
<!-- ``` -->

@@

<!-- How many participants/stimuli are needed for concrete example -->

>>>>

---
title: "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle: "Simulation-Based Sample Size Planning for GLMMs"
author:
  - name: Florian Pargent
    equal-contributor: true
    orcid: 0000-0002-2388-553X
    roles:
      - formal analysis
      - methodology
      - visualization
      - writing
      - editing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
  - name: Timo K. Koch
    corresponding: true
    equal-contributor: true
    orcid: 0000-0001-6728-2063
    email: timo.koch@unisg.ch
    roles:
      - conceptualization
      - formal analysis
      - writing
      - editing
    affiliations:
      - name: University of St. Gallen 
        department: Institute of Behavioral Science & Technology
        address: Torstrasse 25
        city: St. Gallen
        country: Switzerland
        postal-code: 9000
      - ref: lmu
  - name: Anne-Kathrin Kleine
    roles:
      - formal analysis
      - editing
    affiliations:
      - ref: lmu
  - name: Eva Lermer
    roles:
      - conceptualization
      - funding aquisition
      - editing
    affiliations:
      - ref: lmu
      - name: Technical University of Applied Sciences Augsburg
        department: Department of Business Psychology
  - name: Susanne Gaube
    roles:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: lmu
      - name: University College London
        department: Global Business School for Health
author-note:
  disclosures:
    data-sharing: This is version 2.5 of our preprint published at <https://doi.org/10.31234/osf.io/rpjem>. All materials (reproducible manuscript, including R code and all simulation results) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/dhwf4/>. A Quarto Manuscripts website is hosted at <https://Timo-Ko.github.io/glmm_simulation_tutorial/>.
    conflict-of-interest: The authors declare that there were no conflicts of interest with respect to the authorship or the publication of this article.
    financial-support: This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525).
    gratitude: ~
    authorship-agreements: Florian Pargent and Timo K. Koch contributed equally to this work.
abstract: When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. The recent replication crisis underlines the need for more rigorous statistical methodology and adequately powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. However, available methods and software for a priori sample size planning for GLMMs are often limited to specific designs. Tailored data simulation approaches are a more flexible alternative. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori sample size planning with GLMMs. We not only focus on power analysis but also explain how to use the precision of parameter estimates to determine appropriate sample sizes. We conclude with an outlook on the increasing importance of simulation-based sample size planning.
keywords: [power analysis, precision, data simulation, sample size planning, generalized linear mixed model]
bibliography: [Simulation Tutorial Paper.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

```{r}
#| label: setup
#| include: false
library("tidyverse")
library("faux")
library("lme4")
library("multcomp")
library("marginaleffects")
library("tinytable")
library("future")
library("furrr")
library("binom")
library("papaja")
r_refs("r-references.bib")
```

@@

Performing tailored simulation-based sample size planning is more complicated and time-consuming than using the existing software tools outlined in TABLE 1. The most important circumstances under which tailored simulation-based sample size planning is necessary are 1) complex study designs, 2) complex statistical hypotheses, 3) planning for precision, 4) no available prior studies or pilot data.

@@

Although planning for precision [@lakensSampleSizeJustification2022; @maxwellSampleSizePlanning2008; @cummingNewStatisticsWhy2014] has been increasingly discussed as a useful strategy for empirical research, all available software packages for sample size planning with GLMMs are based on power analysis.
Therefore, researchers that want to want to apply an estimation strategy in their studies, without testing any statistical hypotheses, currently cannot use the software packages outlined in TABLE 1.
<!-- As long as one is interested only in confidence intervals of a single model parameter, it might be possible to adapt existing software packages to plan for precision. However, we demonstrate in our case study that in realistic research projects with GLMMs, the contrasts of interests and their respective confidence intervals are usually situated on the scale of the (expected) outcome in contrast to the scale of the model parameters.  -->
However, tailored simulation-based sample size planning can easily handle the planning for precision approach [@maxwellSampleSizePlanning2008]. The only change in procedure is that instead of computing HTs for each simulated dataset and estimating statistical power across repetitions, CIs are computed for each simulated dataset, and the expected width is estimated.

@@

Although the details differ depending on the specific study characteristics, each tailored simulation-based sample size planning requires a series of steps and decisions. We will introduce each step in a theoretical section, followed by the practical application based on a hypothetical case study. All code in this manuscript and simulation results are available in the project’s repository on the Open Science Framework (<https://osf.io/dhwf4/>).

@@

@@

$$
\begin{aligned}
& P(\text{correct diagnosis} | \text{no advice, average student, average scan}) \\
& \quad - P(\text{correct diagnosis} | \text{wrong advice, average student, average scan})
\end{aligned}
$$
For example, the first expression is concerned with the difference between the probability that a correct diagnosis is made if correct AI-advice is presented and the probability that a correct diagnosis is made if no AI-advice is presented. This contrast is quantified for a hypothetical *typical* expert and a *typical* scan, where *typical* is usually defined as an average score on all attributes of the expert or scan.^[Note that a different estimand would be the so-called average treatment effect (ATE). For the ATE, the probability contrast is defined for each combination of expert and scan, and then these contrasts are averaged across all experts and scans from the target population [@lundbergWhatYourEstimand2021].].
To complete our definition of the estimand, we have to define our target population that consists of persons, stimuli, and AI-advice: With respect to persons, we are only interested in expert radiologists and medical students at German universities. With respect to stimuli, we are only interested in the head CT scans made from subjects that do or do not suffer from intracerebral hemorrhage. Lastly, we are only interested in AI-advice given by a specific AI-enabled diagnostic decision support system.

@@

The discussed inequality of conditional and marginal effects in GLMMs [@fahrmeirRegressionModelsMethods2021] makes their interpretation more difficult.
One must be careful when specifying parameter values based on previous studies or pilot data that use the marginal interpretation (e.g., a pilot study providing an estimate of how often radiologists make an accurate diagnosis based on brain scans). 
However, this does not mean that we cannot use the marginal interpretation (average probability across persons and items) to inform plausible parameter values: When parameter values have been selected, we can compute the implied marginal distributions and compare this information to our domain knowledge.
Then, we can iteratively adjust the parameter values until we are satisfied with the implied distributions.
In the last section, we simulated a large dataset and computed descriptive statistics, the relative frequencies of correct diagnoses, for each experimental condition.
We will now use the model implied probability of each simulated data point (stored in the variable `y_prob`) to visualize the whole model implied marginal distribution of correct diagnoses for each experimental condition.

```{r}
#| label: fig-margdist1
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions."
#| echo: true
library(ggdist)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
"no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
"no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```
@fig-margdist1 shows the model implied marginal distributions, including the mean, 66% and 95% intervals. We can see that, indeed, the average probabilities (black dots) slightly differ from the probabilities of average subjects and items considered in the previous section. This difference increases with the variability of the random effects.
We can use plots like the one above as a useful tool to decide whether the specified standard deviations of the subject and item random intercepts ($\sigma_S$ and $\sigma_I$) are reasonable by comparing the ranges and overlap between conditions to domain knowledge.

```{r}
#| label: fig-margdist2
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of item random intercepts to 0.01."
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0i = 0.01)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
"no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
"no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```
In the next plot, we have set the item standard deviation to almost zero ($\sigma_I = 0.01$). This gives us a better way to see the variability between persons. As an example, @fig-margdist2 reveals a number of implicit assumptions about the comparison between experts and non-experts: With wrong advice, virtually all experts have a higher probability of making a correct diagnosis compared to non-experts when considering only items with average difficulty. In contrast, there is considerable overlap in probability between experts and non-experts with no advice and even higher overlap with correct advice. Patterns like these should be considered carefully and discussed with the domain experts. Parameter values ($\beta$ parameters, and $\sigma_S$) should be adjusted if the implications do not seem reasonable.
We could also have a closer look at variability between items by setting the subject standard deviation to almost zero ($\sigma_S = 0.01$).
```{r}
#| include: false
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 0.01)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
"no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
"no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
"no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

The final plot demonstrates that these plots are also useful for spotting standard deviations that are specified too high. For @fig-margdist3 we have set $\sigma_S = 3$ and $\sigma_I = 3$. This implies that in each experimental condition, the probabilities that a subject solves an item are usually close to either 0 or 1, which is not a plausible assumption. These high standard deviations do not account for the inherent variability and complexity of human performance. For example, we would expect that a participant with low ability compared to other task experts to solve a difficult item with a probability substantially larger than zero even when presented with wrong advice.
```{r}
#| label: fig-margdist3
#| echo: false
#| fig-cap: "Marginal distributions including means, 66% and 95% confidence intervals for all experimental conditions while setting the standard deviation of subject and item random intercepts to 3."
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 3, sd_u0i = 3)
dat |> 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) |>
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) |> 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45",
    breaks = "Sturges") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@@

In our case study, we use the lme4 R package [@batesFittingLinearMixedEffects2015], which is a state-of-the-art tool for fitting frequentist GLMMs.^[For Bayesian GLMMs, the brms R package is currently the most prominent option [@burknerBrmsPackageBayesian2017].]
For the current example, we simulate data according to our model, in which 100 subjects respond to 50 items (we use `set.seed` to make the simulation reproducible). However, for the sake of the exercise, we can imagine that this would be real data resulting from our future experiment and think about how we would analyze this data.

@@

We could use our knowledge of the structure of the GLMM to determine the exact formula needed to compute the contrasts of interest and then plug in the parameter estimates manually from the `summary(fit)` output.
However, this would be tedious and we can use R to compute this contrast without doing the math.
Using the first contrast *(correct advice, expert) - (no advice, expert)* as our example, we could apply the `predict` function of the lme4 package to compute the predicted probability for a correct diagnosis based on our fitted model, plug in the two sets of predictor values, and compute the difference between the two estimated probabilities.

```{r}
#| echo: true
grid1 <- data.frame(advice_present = c(1, 0), advice_correct = c(1, 0), 
  expert = c(1, 1))
grid1
pred <- predict(fit, newdata = grid1, type = "response", re.form = NA)
pred
pred[1] - pred[2]
```

The argument `type = "response"` specifies that predictions are made on the probability scale (instead of the log-odds scale of the $\beta$ parameters), while `re.form = NA` sets all random effects to 0.
We could use this method to compute point estimates for all four contrasts that are part of our estimand.
However, depending on whether we are interested in hypothesis testing or parameter estimation, we also need a method to compute HTs or CI.
The marginaleffects package [@R-marginaleffects] is a very flexible, increasingly popular package to compute HTs and CIs for contrasts with a variety of statistical models, including GLMMs estimated with lme4.
First, we specify a grid of all combinations of predictor variable and then compute estimated probabilities for all experimental conditions in our experiment with the `predictions` function:

```{r}
#| echo: true
library(tidyverse)
library(marginaleffects)
library(tinytable)
grid2 <- expand_grid(advice_present = 0:1, 
  advice_correct = 0:1, expert = 0:1)
grid2
preds <- predictions(fit, newdata = grid2, 
  type = "response", re.form = NA)
print(preds, style = "tinytable") |> theme_tt(theme = "resize")
```

<!-- print(probs, style = "tinytable") -->

@@

```{r}
#| echo: true
#| size: footnotesize
contrasts <- preds |> 
  hypotheses(hypothesis = c(
    "b8 = b2",  # (correct advice, expert) - (no advice, expert)
    "b2 = b6",  # (no advice, expert) - (wrong advice, expert) 
    "b7 = b1",  # (correct advice, no expert) - (no advice, no expert)
    "b1 = b5"), # (no advice, no expert) - (wrong advice, no expert)
    equivalence = c(0, 0))
print(contrasts, style = "tinytable") |> theme_tt(theme = "resize")
```

@@

If we chose hypothesis testing for our case study, we would test a combined null hypothesis $H_0$ that consists of four separate null hypotheses:
$$
\begin{aligned}
H_{01}:\ & P(Y=1|advice\_present = 1, advice\_correct = 1, expert = 1, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \\
H_{02}:\ &P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 1, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \\
H_{03}:\ &P(Y=1|advice\_present = 1, advice\_correct = 1, expert = 0, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) \\
H_{04}:\ & P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 1, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0)
\end{aligned}
$$
The combined null hypothesis $H_0$ should only be rejected if **all** individual null hypotheses are rejected [i.e., intersection-union setting\; @dmitrienkoTraditionalMultiplicityAdjustment2013].
In such cases, the error probabilities do not accumulate, and we would waste power when correcting for multiple tests.

@@

@@
